EuPPollNet: A European database of plant-pollinator networks

### 2024-05-22 Manuscript/EuPPollNet_Ms.qmd
edits

To get a general overview of some of the main aspects of these set of plant-pollinator networks, we evaluated how connectance and nestedness change across the latitudinal range of studies and bioclimatic regions. We implemented linear models, with the network metrics as the response variable, and both latitude and bioclimatic region as predictors in two separate models. Finally, with the help of a Tukey test from the package _emmeans_, we evaluated if nestedness and connectance differ statistically across bioclimatic regions. In addition, we explored the association between connectance and nestedness with the number of species per network (i.e., geometric mean of plants and pollinators) by conducting Kendall rank correlation coefficient. Finally, we evaluated if the different networks are more or less nested than expected by chance. Importantly, we implemented two different metrics of nestedness given the different analytical objectives. Common metrics of nestedness are highly correlated to network size, limiting our ability to compare this metric across networks. To avoid this, we implemented a normalised version of nestedness (NODFc) that corrects by connectance and the number of species [@song2017]. This metric was calculated with the help of the **maxnodf** package [@hoeppke2021], which employs a computational efficient approach to calculate it. However, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008, as this approach allow us to compare our results with previous published nestedness analyses in plant-pollinator networks and only compares each unique network against their randomized versions. We calculated 100 null models for each network with the help of the _vaznull_ function from the package **Bipartite** [@dormann2008] that implements the null model from @vazquez2005. These null networks have the same connectance and number of plant and pollinator species as the empirical ones, but different marginal totals. In other words, interactions are reshuffled with almost equal probabilities to their relative abundances. Connectance was estimated for each network by using the function _networklevel_ from **Bipartite**.

@@

Connectance and nestedness did not show relevant trends across the latitudinal range of studies (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ = `r round(connectance_lat_pval, 2)`, **Figure 6a**; NODFc: R$^2$ = `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6b**). However, differences were found for connectance but not nestedness across bioclimatic regions. Specifically, the alpine region differed statistically in connectance to the other bioclimatic regions (Tukey test _P_ < 0.01). In addition, the atlantic and continental bioclimatic regions also showed statistical meaningful differences in connectance (Tukey test _P_ < 0.01). Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and as expected, followed an negative exponential relationship with the geometric mean of plant and pollinator species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6c**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and as expected for this metric, were independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S3**). Finally, we found that only `r round(over_z_percent, 2)`% of networks were statistically more nested than expected by chance, with `r round(no_diff_z_percent, 2)`% showing no statistical difference, and none being less nested than null expectations (**Figure 6d**).

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 6}. Graphs (a-b) indicate network connectance and nestedness across the latitudinal range of the studies from the database. In addition, the bioclimatic region where each network is located is indicated with points of different shapes and colours. Graph (c) shows the association between network connectance and the geometric mean of plant and pollinator species per network. Graph (d) shows the distribution of z-scores when comparing the nestedness from the empirical networks with their randomised counterparts (100 null models for each network). The vertical red dashed lines represent the z critical value for a two tailed test with alpha = 0.05. Z-scores to the left of the first vertical red dashed line indicate that networks are less nested than expected by chance (red), those between the two dashed lines indicate no statistical difference from random expectations (green), and those to the right indicate that networks are more nested than expected by chance (blue).", fig.width=10, fig.height=7, fig.align='center'}

# Load libraries

library(psych)
# Load data
data = readRDS("../Data/3_Final_data/Interaction_data.rds")
metrics_by_network = readRDS("../Data/Working_files/metrics_by_network.rds")
null_networks = readRDS("../Data/Working_files/null_networks.rds")
null_metrics_networks = readRDS("../Data/Working_files/null_metrics_networks.rds")

# Prepare data
data = data %>%
dplyr::mutate(Year = lubridate::year(Date), 
              Month = lubridate::month(Date), 
              Day = lubridate::day(Date))

# Subset data for studies with only bumblebees
v = c("13_Karise", "22_Kallnik")
subsetting = unique(data$Study_id)

long_format = data %>% 
mutate(Network_id = paste0(Study_id, "_", Network_id, "_", Year)) %>% 
mutate(Network_id = str_replace_all(Network_id, " ","_")) %>% 
filter(!Study_id %in% v) %>% 
select(Network_id, Plant_accepted_name, Pollinator_accepted_name)

# Filter networks larger than 4 plants and 4 pollinator species
v2 = long_format %>% 
group_by(Network_id) %>% 
summarise(Poll_spp = n_distinct(Plant_accepted_name),
          Plant_spp = n_distinct(Pollinator_accepted_name)) %>% 
ungroup() %>% 
filter(Poll_spp >= 4 & Plant_spp >= 4) %>% 
pull(Network_id)

long_format = long_format %>% 
filter(Network_id %in% v2)

# Filter networks with more links than species
links = long_format %>% 
group_by(Network_id, Plant_accepted_name, Pollinator_accepted_name) %>%
distinct() %>% 
group_by(Network_id) %>% 
summarise(links = length(Network_id))

spp = long_format %>% 
group_by(Network_id) %>% 
summarise(n_plants = n_distinct(Plant_accepted_name),
          n_polls = n_distinct(Pollinator_accepted_name)) %>% 
mutate(n_total = n_plants + n_polls)

vector_to_include = left_join(links, spp)

v3 = vector_to_include %>% 
filter(links > n_total) %>% 
pull(Network_id)

long_format = long_format %>% 
filter(Network_id %in% v3)

# Explore how NODFc changes across the Y-X AXES (latitude-Longitude)
long_format_coords = data %>% 
mutate(Network_id = paste0(Study_id, "_", Network_id, "_", Year)) %>% 
mutate(Network_id = str_replace_all(Network_id, " ","_")) %>% 
select(Network_id, Latitude, Longitude, Bioregion) %>% 
distinct()

# Merge data
metrics_by_network_coords = left_join(metrics_by_network, long_format_coords)

# Connectance plots
# 1st Latitude
p1 = metrics_by_network_coords %>% 
ggplot(aes(Latitude, Connectance)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.15, size=2.5, alpha=1) +
theme_bw() +
ylab("Connectance") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
))+
ggtitle("(a)")

# 2nd Plot
p2 = metrics_by_network_coords %>% 
ggplot(aes(Latitude, Normalised_nestedness)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.15, size=2.5, alpha=1) +
theme_bw() +
ylab("NODFc") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
ggtitle("(b)")

# Merge and process data
d = left_join(metrics_by_network, null_metrics_networks)
d = d %>% 
mutate(z_score = (Classic_nestedness - Mean_null_classic_nestedness) / Deviation_null_classic_nestedness)

p = 0.05 #cutoff probability 95% confidence
critical_value = qnorm(p/2) #double tail probability divide by 2

d1 = d %>%
mutate(infra_over_represented = case_when(
z_score < -abs(critical_value) ~ "Under-represented",
between(z_score, -abs(critical_value), abs(critical_value)) ~ "No statistical difference",
    z_score > abs(critical_value) ~ "Over-represented"
  ))

# 2nd Species (geometric mean)
species_number = long_format %>% 
group_by(Network_id) %>% 
summarise(N_plants = n_distinct(Plant_accepted_name),
          N_pollinators = n_distinct(Pollinator_accepted_name)) %>% 
group_by(Network_id) %>% 
summarise(geometric_mean_spp = geometric.mean(c(N_plants, N_pollinators)))

metrics_by_network_spp_number = left_join(metrics_by_network, species_number)

p3 = metrics_by_network_spp_number %>% 
ggplot(aes(geometric_mean_spp, Connectance)) +
geom_point(pch = 21, fill = "azure3", size=2.5, stroke=0.25, alpha=1) +
theme_bw() +
xlab("Species")+
ggtitle("(c)")

# Plot
p4 = d1 %>% 
ggplot(aes(x = z_score)) +
geom_histogram(aes(y = ..density.., fill = infra_over_represented), 
               bins = 100, alpha = 0.5, color = "black", position = "stack", size=0.1) +
stat_function(data = d1, fun = function(x) dnorm(x, mean = mean(d1$z_score), sd = sd(d1$z_score)) * 1.7, 
                n = 1000, inherit.aes = FALSE, color = "gray18", size = 1) +
theme_bw() +
coord_cartesian(expand = FALSE) +
geom_vline(xintercept = -abs(critical_value), linetype = "longdash", colour = "red")+
geom_vline(xintercept = abs(critical_value), linetype = "longdash", colour = "red") +
ylab("Density") +
xlab("Z scores (NODF)")  +
xlim(-6, 9) +
ylim(0, 0.8) +
scale_fill_manual(name = "Observed NODF against null",
                    limits = c("Under-represented", "No statistical difference", "Over-represented"),
                    labels = c("Less nested", "No difference", "More nested"),
                    values = c("Under-represented" = "coral2", 
                               "No statistical difference" = "palegreen3", 
                               "Over-represented" = "cyan3")) +
ggtitle("(d)")

# Layout plots
a = p1 + p2 + plot_layout(guides = 'collect') & theme(legend.position = "right",
                                                      legend.justification = "left")
b = (p3 + p4)  & theme(legend.justification = "left")
a / b  & theme(axis.title =  element_text(size=16, face="bold"),
               plot.title = element_text(size=22),
               panel.border = element_rect(colour = "black",
               fill=NA, size=1),
               axis.text=element_text(size=12),
               legend.text = element_text(size=10),
               legend.key.size = unit(6, "mm"),
               legend.spacing.y = unit(0.5, "cm"),
               legend.box.spacing = unit(0, "pt"),
               legend.title = element_text(face = "bold"))

@@

>>>>

To provide a general overview of the main aspects of this set of plant-pollinator networks, we evaluated network patterns of connectance and nestedness across the latitudinal range of studies and bioclimatic regions. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. However, these network metrics are not independent of sampling effort (e.g., number of species) and to allow comparisons across them, we implemented “standardised” versions of connectance and nestedness. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species  (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. This was done with the help of a Beta regression implemented with the package **betareg** [@cribari2010]. Second, we implemented a metric of nedstedness (i.e., NODFc) that allows the comparison across networks as it corrects by connectance and the number of species [@song2017]. This metric was calculated with the help of the **maxnodf** package [@hoeppke2021], which employs a computational eﬀicient approach to calculate it. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we explored the association between connectance and nestedness with the number of species per network by conducting Kendall rank correlation coefficient in order to be able to compare the strength of this association across both network metrics.

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008, as this approach allow us to compare our results with previous published nestedness analyses in plant-pollinator networks and only compares each unique network against their randomized versions. We calculated 100 null models for each network with the help of the vaznull function from the package **bipartite** [@dormann2008] that implements the null model from @vazquez2005. These null networks have the same connectance and number of plant and pollinator species as the empirical ones, but different marginal totals. Both connectance and nestedness (NODF) were estimated for each network by using the function _networklevel_ from **bipartite**.

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and as expected, followed an negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and as expected for this metric, were independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S3**). We found that only `r round(over_z_percent, 2)`% of networks were statistically more nested than expected by chance, with `r round(no_diff_z_percent, 2)`% showing no statistical difference, and none being less nested than null expectations (**Figure 6b**). Latitude explained little of the observed variability of residual connectance and nestedness across networks (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ = `r round(connectance_lat_pval, 2)`, **Figure 6a**; NODFc: R$^2$ = `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6b**). Overall, networks towards higher latitudes showed lower residual connectance and higher nestedness that networks located in lower latitudes.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 6}. Graph (a) shows the association between network connectance and the geometric mean of plant and pollinator species per network on a log-scale with the respective fitted line from a Beta regression. Graph (b) shows the distribution of z-scores when comparing the nestedness from the empirical networks with their randomised counterparts (100 null models for each network). The vertical red dashed lines represent the z critical value for a two tailed test with alpha = 0.05. Z-scores to the left of the first vertical red dashed line indicate that networks are less nested than expected by chance (red), those between the two dashed lines indicate no statistical difference from random expectations (green), and those to the right indicate that networks are more nested than expected by chance (blue). Graphs (c-d) show the fitted regression of residual connectance and nestedness across the latitudinal range of the studies from the database. In addition, the bioclimatic region of each network is indicated with points of different shapes and colours.", fig.width=10, fig.height=7, fig.align='center'}

# Load libraries
library(dplyr)
library(ggplot2)
library(stringr)
library(patchwork)

# Load plots generated in scripts 7_2_Nestedness and 7_2_Connectance

connectance_species = readRDS("../Data/Manuscript_info/Figure6_connectance_species.rds")
connectance_latitude = readRDS("../Data/Manuscript_info/Figure6_connectance_latitude.rds")
nestedness_z_scores = readRDS("../Data/Manuscript_info/nestedness_z_scores.rds")
nestedness_latitude = readRDS("../Data/Manuscript_info/nestedness_latitude.rds")

#Prepare first connectance plots
library(betareg)
model = betareg(Connectance ~ log_geometric_mean_spp, data = connectance_species)
p1 = ggplot(connectance_species, aes(x = log_geometric_mean_spp, y = Connectance))+
geom_point(pch = 21, fill = "azure3", size=2.5) +
geom_line(aes(y = predict(model, connectance_species)), colour= "black")+
xlab("log(Species)") +
theme_bw() +
ggtitle("(a)")

p3 = connectance_latitude %>% 
ggplot(aes(Latitude, residual_conectance)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.25, size=2.5) +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
geom_line(data = connectance_latitude,aes(x=Latitude, y = fit), inherit.aes = FALSE)+
geom_ribbon(data = connectance_latitude, 
    aes(x=Latitude,ymin = lwr, ymax = upr), inherit.aes = FALSE,
    fill = "grey70", alpha= 0.3) +
theme_bw()+
ylab("Residual connectance") +
ggtitle("(c)")

#Prepare now nestedness plots
# Plot
p = 0.05 #cutoff probability 95% confidence
critical_value = qnorm(p/2) #double tail probability divide by 2

nestedness_z_scores = nestedness_z_scores %>%
mutate(infra_over_represented = case_when(
z_score < -abs(critical_value) ~ "Under-represented",
between(z_score, -abs(critical_value), abs(critical_value)) ~ "No statistical difference",
    z_score > abs(critical_value) ~ "Over-represented"
  ))

p2 = nestedness_z_scores %>% 
ggplot(aes(x = z_score)) +
geom_histogram(aes(y = ..density.., fill = infra_over_represented), 
               bins = 100, alpha = 0.5, color = "black", position = "stack") +
stat_function(data = nestedness_z_scores, fun = function(x) dnorm(x, mean = mean(nestedness_z_scores$z_score), sd = sd(nestedness_z_scores$z_score)) * 1.7, 
              n = 1000, inherit.aes = FALSE, color = "gray18", size = 1) +
theme_bw() +
coord_cartesian(expand = FALSE) +
geom_vline(xintercept = -abs(critical_value), linetype = "longdash", colour = "red") +
geom_vline(xintercept = abs(critical_value), linetype = "longdash", colour = "red") +
ylab("Density") +
xlab("Z scores (NODF)")  +
xlim(-6, 9) +
ylim(0, 0.8) +
scale_fill_manual(name = "Observed against null",
                  limits = c("Under-represented", "No statistical difference", "Over-represented"),
                  labels = c("Less nested", "No difference", "More nested"),
                  values = c("Under-represented" = "coral2", 
                             "No statistical difference" = "palegreen3", 
                             "Over-represented" = "cyan3")) +
ggtitle("(b)")

# 2nd Plot
p4 = nestedness_latitude %>% 
ggplot(aes(Latitude, Normalised_nestedness)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.25, size=2.5) +
theme_bw() +
ylab("NODFc") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
geom_line(data = nestedness_latitude,aes(x=Latitude, y = fit), inherit.aes = FALSE)+
geom_ribbon(data = nestedness_latitude, 
    aes(x=Latitude,ymin = lwr, ymax = upr), inherit.aes = FALSE,
    fill = "grey70", alpha= 0.3) +
theme_bw()+
ggtitle("(d)")

# Layout plots
#a = (p1 + p2) 
#b = p3 + p4 + plot_layout(guides = 'collect') & theme(legend.position = "right")
#a / b

# Layout plots
a = (p1 + p2)  & theme(legend.position = "right", legend.justification = "left")
b = p3 + p4 + plot_layout(guides = 'collect') & theme(legend.justification = "left")
a / b  & theme(axis.title =  element_text(size=16, face="bold"),
               plot.title = element_text(size=22),
               panel.border = element_rect(colour = "black",
               fill=NA, size=1),
               axis.text=element_text(size=12))

@@

metrics_by_network_spp_number = readRDS("../Data/Working_files/metrics_by_network_spp_number.rds")
